Research Proposal

### 2012-03-13 dissertation-proposal/research-proposal.tex
More changes for second draft. TODO method

\begin{namelist}{xxxxxxxxxxxx}
\item[{\bf Title:}]
	Optional Static Type Checking for Dynamic Languages
\item[{\bf Author:}]
	Ambrose Bonnaire-Sergeant
\item[{\bf Supervisor:}]
	Rowan Davies
\end{namelist}

@@

Dynamic languages are designed to be convenient for the programmer to
write programs quickly and aspire to get out of the programmer's
way as much as possible. When programs grow large and stabilize,
the features of static languages are missed, specifically static
type checking.

@@

A recent example is Dart 
\cite{Dart:2012:Online}, a dynamically typed language that allows optional
static checks.

Good type systems deeply integrate with the language they are written for.
Statically typed languages usually include a type system in early releases,
which is able to evolve with the language while it is malleable.

The type system understands common idioms
and becomes the programmer's ally.

The tradeoffs of both styles has been subject to much debate, but some
points have general agreement. Dynamically typed languages
are generally more convenient for writing smaller 
programs, whereas statically typed languages are usually preferred 
when writing large programs.

Typed Racket \cite{TypedScheme:2010}
, which provides a statically typed variation of Racket 
with the intention of safely porting untyped code to typed code. 
Typed code can interact with untyped code while preserving some properties
of typed code by, for example, delaying static type checks until runtime.
Porting untyped Racket code to Typed Racket generally involves 
adding type annotations to top level declarations with minimal (usually zero)
modification to existing code structure.

Type checking is often an operation complected with compilation. This is
because languages that have type checkers generally enforce the rule that
if a program does not satisfy the type checker, then the program should not be 
allowed to compile. When considering an optional type checker, it makes sense
to provide type checking as a simple operation, callable a la carte. 
This relieves a common restriction associated with statically typed languages;
programs can now compile regardless of the results of the 
type checker.

@@

My goal is to develop an optional static type system for Clojure, 
intended for practical use, to be provided as a library.

It will be based on the lessons learnt throught the development
of Typed Scheme, and as a response to Tobin-Hochstadt's \cite{TypedScheme:2010}
suggestion to add type systems to dynamic languages.

I expect this library to be an important asset to Clojure's feature set.
I expect it will become a factor in language decision and justification for programming projects.

>>>>

\begin{namelist}{xxxxxxxxxxxx}
\item[{\bf Title:}]
	Optional Static Type Checking for Clojure
\item[{\bf Author:}]
	Ambrose Bonnaire-Sergeant
\item[{\bf Supervisor:}]
	Rowan Davies
\end{namelist}

@@

Dynamic languages are designed to be convenient for
writing programs quickly, and aspire to get out of the programmer's
way as much as possible. When programs grow large and stabilize,
some features of static languages are missed, specifically static
type checking.

@@

Clojure is a dynamic, functional language with implementations for the Java Virtual
Machine and Common Language Runtime, and compiling to Javascript. Clojure
is also a Lisp, which makes it a good candidate to test the ideas developed in
Typed Scheme \cite{TypedScheme:2010}.

@@

My goal is to develop a prototype optional static type system for Clojure, 
eventually intended for practical use, to be provided as a library.

It will be based on the lessons learnt throught the development
of Typed Scheme, and as a response to Tobin-Hochstadt's \cite{TypedScheme:2010}
suggestion to add type systems to existing dynamic languages.

There are several challenges to creating a satisfactory type system for Clojure.

Multimethods play a significant role in Clojure, and a satisfactory
type system for Clojure should understand them to some degree.
There is some experience statically typing multimethods
\cite{Millstein02modularstatically}
, but this is a challenge and will probably not be completed in the
timeframe proposed.

Clojure's core library includes variable-arity functions which
are not easily typed with current static type systems. 
Providing satisfactory types for functions 
like Clojure's 'map', 'filter', and 'reduce' require
support for non-uniform variable arity polymorphism. 
Strickland, Tobin-Hochstadt and Felleisen \cite{Strickland:2009:PVP:1532974.1532978}.
describe their approach for non-uniform variable arity polymorphism, as used
in Typed Racket. It is not clear if including a similar system is 
achievable in the given timeframe, but would be a very useful feature.

Occurence typing is type checking technique developed for Typed Scheme 
\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
and improved for Typed Racket
\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}.
It helps the type checker understand common programming idioms 
with minimal type annotations.
The general approach of occurence typing appears to fit Clojure well,
but the implementation challenges are not clear. Occurence typing,
or something similar, would be essential to a practical system.

Ensuring vigorous type safety is an important aspect of Typed Scheme 
\cite{TypedScheme:2010} ,
especially when interacting between untyped and typed modules.
I do not expect to concentrate on every combination of cross-module
interaction. In particular, safely using typed code from untyped code
is not important for my needs.
